\bf Title here

### 2020-08-06 paper/paper.tex
move changes made by Nicolas to Rmd

\bigskip
\begin{abstract}
Friedman \& Tukey commented on their initial paper on projection pursuit
in 1974 that ``the technique used for maximising the projection index
strongly influences both the statistical and the computational aspects
of the procedure.'' While many projection pursuit indices have been
proposed in the literature, few concerns the optimisation procedure. In
this paper, we developed a system of diagnostics aiming to visually
learn how the optimisation procedures find its way towards the optimum.
This diagnostic system can be applied more generally to help
practitioner to unveil the black-box in randomised iterative
(optimisation) algorithms. An R package, ferrn, has been created to
implement this diagnostic system.
\end{abstract}

@@

Below are listed several issues in projection pursuit optimisation. Some are
general optimisation problems, while others are more specific to PP
optimisation.

\begin{itemize}
\item
  \emph{Finding global maximum}: Although finding local maximum is
  relatively easy with developed algorithms, it is generally hard to
  guarantee global maximum in a problem where the objective function is
  complex or the number of decision variables is large. Also, there are
  discussions on how to avoid getting trapped in a local optimum in the
  literature.
\item
  \emph{optimising non-smooth function}: When the objective function is
  non-differentiable, derivative information can not be obtained, which
  means traditional gradient- or Hessian- based methods are not
  feasible. Stochastic optimisation method could be an alternative to
  solve these problems.
\item
  \emph{computation speed}: The optimisation procedure needs to be fast
  to compute since tours produce real-time animation of the projected
  data.
\item
  \emph{consistency result in stochastic optimisation}: In stochastic
  algorithm, researchers usually set a seed to ensure the algorithm
  produces the same result for every run. This practice supports
  reproducibility, while less efforts has been made to guarantee
  different seeds will provide the same result.
\item
  \emph{high-dimensional decision variable}: In projection pursuit, the
  decision variable includes all the entries in the projection matrix,
  which is high-dimensional. Researchers would be better off if they could
  understand the relative position of different projection matrix in the
  high-dimensional space.
\item
  \emph{role of interpolation in PP optimisation}: An optimisation
  procedure usually involves iteratively finding projection bases that
  maximise the index function, while tour requires geodesic
  interpolation between these bases to produce a continuous view for the
  users. It would be interesting to see if the interpolated bases could,
  in reverse, help the optimisation reach faster convergence.
\end{itemize}

@@

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated} is a
non-derivative procedure based on a non-increasing cooling scheme
\(T(l)\). Given an initial \(T_0\), the temperature at iteration \(l\)
is defined as \(T(l) = \frac{T_0}{\log(l + 1)}\). The simulated
annealing algorithm works as follows. Given a neighbourhood parameter
\(\alpha\) and a randomly generated orthonormal basis \(\mathbf{A}_{\text{rand}}\), a
candidate basis is sampled by
\(\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{\text{rand}}\).
If the index value of the candidate basis is larger than the one of the
current basis, the candidate basis is outputted. While if it is smaller,
the candidate has another chance of being accepted with probability
\(P= \min\left\{\exp\left[\frac{I_{cur} - I_{l}}{T(l)}\right],1\right\}\)
where \(I_{\cdot}\) denotes the index value of a given basis. The
algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

@@

We simulate some random variables of size 1000 with different
structures. \texttt{x1}, \texttt{x8}, \texttt{x9} and \texttt{x10} are
simulated from normal distribution with zero mean and variance of one as
in equation \eqref{eq:sim-norm}. When using projection pursuit to explore
the data structure based on its departure from normality, the entry in
the projection basis for these variables should be close to zero in
theory. \texttt{x2} to \texttt{x7} are mixture of normal distributions
with different weights and locations. Equation \eqref{eq:sim-x2} to
\eqref{eq:sim-x7} outlines the distribution where each variable is
simulated from and Figure \ref{origin-data} shows the histogram of each
variable except \texttt{x3}. All the variables are then scaled to have
unit variance before running the projection pursuit.

@@

We use the same dataset as the toy example above to explore the search
function \texttt{search\_better} and we want to learn how the index
value changes on the interpolation path for the \texttt{holes} index. From
the left panel of Figure \ref{interruption}, we observe that when
interpolating from the current basis to the target basis, the index
value may not be monotone: we could reach a basis with higher index
value than the target basis on the interpolation path. In this sense, we
would be better off using the basis with the highest index value on the
interpolation path as the current basis for the next iteration (rather
than using the target basis). Hence, an interruption is constructed to
accept the interpolating bases only up to the one with the largest index
value. After implementing this interruption, the search finds a higher
final index value with fewer steps as shown in the right panel of Figure
\ref{interruption}.

@@

\[\max \left[F_{\mathbf{P}}(u) - F_{y}(u)\right]\]
can be seen as a function of the projection matrix
\(\mathbf{A}_{p \times 1}\) and hence a valid index function.

@@

Using different seeds in \texttt{search\_better} allows us to find both
local maxima as in Figure \ref{1d-2var-different-seeds}. Comparing the
maximum of both, we can see that the global maximum happens when
\texttt{x2} is found. It is natural to ask then if there is an algorithm
that can find the global maximum without trying on different seeds?
\texttt{search\_better\_random} manages to do it via a
Metropolis-Hasting random search as shown in Figure
\ref{1d-2var-better-random}, although at a higher cost of number of
points to evaluate.

>>>>

\bigskip
\begin{abstract}
Friedman \& Tukey commented on their initial paper on projection pursuit
in 1974 that ``the technique used for maximising the projection index
strongly influences both the statistical and the computational aspects
of the procedure.'' While many projection pursuit indices have been
proposed in the literature, few concerns the optimisation procedure. In
this paper, we developed a system of diagnostics aiming to visually
learn how the optimisation procedures find its way towards the optimum.
This diagnostic system can be applied to more generally to help
practitioner to unveil the black-box in randomised iterative
(optimisation) algorithms. An R package, ferrn, has been created to
implement this diagnostic system.
\end{abstract}

@@

Below are listed several issues in projection pursuit optimisation. Some
are general optimisation problems, while others are more specific too PP
optimisation.

\begin{itemize}
\item
  \emph{Finding global maximum}: Although finding local maximum is
  relatively easy with developed algorithms, it is generally hard to
  guarantee global maximum in a problem where the objective function is
  complex or the number of decision variables is large. Also, there are
  discussions on how to avoid getting trapped in a local optimum in the
  literature.
\item
  \emph{optimising non-smooth function}: When the objective function is
  non-differentiable, derivative information can not be obtained, which
  means traditional gradient- or Hessian- based methods are not
  feasible. Stochastic optimisation method could be an alternative to
  solve these problems.
\item
  \emph{computation speed}: The optimisation procedure needs to be fast
  to compute since the tour produces real-time animation of the
  projected data.
\item
  \emph{consistency result in stochastic optimisation}: In stochastic
  algorithm, researchers usually set a seed to ensure the algorithm
  produces the same result for every run. This practice supports
  reproducibility, while less efforts has been made to guarantee
  different seeds will provide the same result.
\item
  \emph{high-dimensional decision variable}: In projection pursuit, the
  decision variable includes all the entries in the projection matrix,
  which is high-dimensional. Researchers would be better off if they
  could understand the relative position of different projection matrix
  in the high-dimensional space.
\item
  \emph{role of interpolation in PP optimisation}: An optimisation
  procedure usually involves iteratively finding projection bases that
  maximise the index function, while tour requires geodesic
  interpolation between these bases to produce a continuous view for the
  users. It would be interesting to see if the interpolated bases could,
  in reverse, help the optimisation reach faster convergence.
\end{itemize}

@@

Simulated annealing
\citep[\citet{kirkpatrick1983optimization}]{bertsimas1993simulated} is a
non-derivative procedure based on a non-increasing cooling scheme
\(T(l)\). Given an initial \(T_0\), the temperature at iteration \(l\)
is defined as \(T(l) = \frac{T_0}{\log(l + 1)}\). The simulated
annealing algorithm works as follows. Given a neighbourhood parameter
\(\alpha\) and a randomly generated orthonormal basis
\(A_{\text{rand}}\), a candidate basis is sampled by
\(\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{\text{rand}}\).
If the index value of the candidate basis is larger than the one of the
current basis, the candidate basis is outputted. While if it is smaller,
the candidate has another chance of being accepted with probability
\(P= \min\left\{\exp\left[\frac{I_{cur} - I_{l}}{T(l)}\right],1\right\}\)
where \(I_{\cdot}\) denotes the index value of a given basis. The
algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

@@

We simulate some random variables of size 1000 with different
structures. \texttt{x1}, \texttt{x8}, \texttt{x9} and \texttt{x10} are
simulated from normal distribution with zero mean and variance of one as
in equation \ref{eq:sim-norm}. When using projection pursuit to explore
the data structure based on its departure from normality, the entry in
the projection basis for these variables should be close to zero in
theory. \texttt{x2} to \texttt{x7} are mixture of normal distributions
with different weights and locations. Equation \ref{eq:sim-x2} to
\ref{eq:sim-x7} outlines the distribution where each variable is
simulated from and Figure \ref{origin-data} shows the histogram of each
variable except \texttt{x3}. All the variables are then scaled to have
unit variance before running the projection pursuit.

@@

We use the same dataset as the toy example above to explore the search
function \texttt{search\_better} and we want to learn how the index
value changes on the interpolation path for the \texttt{holes} index.
From the left panel of Figure \ref{interruption}, we observe that when
interpolating from the current basis to the target basis, the index
value may not be monotone: we could reach a basis with a higher index
value than the target basis on the interpolation path. In this sense, we
would be better off using the basis with the highest index value on the
interpolation path as the current basis for the next iteration (rather
than using the target basis). Hence, an interruption is constructed to
accept the interpolating bases only up to the one with the largest index
value. After implementing this interruption, the search finds higher
final index value with fewer steps as shown in the right panel of Figure
\ref{interruption}.

@@

\[\max \left[F_{\mathbf{P}}(u) - F_{y}(u)\right]\]

can be seen as a function of the projection matrix
\(\mathbf{A}_{p \times 1}\) and hence a valid index function.

@@

Using different seeds in \texttt{search\_better} allows us to find both
local maxima d as in Figure \ref{1d-2var-different-seeds}. Comparing the
maximum of both, we can see that the global maximum happens when
\texttt{x2} is found. It is natural to ask then if there is an algorithm
that can find the global maximum without trying on different seeds?
\texttt{search\_better\_random} manages to do it via a
Metropolis-hasting random search as shown in Figure
\ref{1d-2var-better-random}, although at a higher cost of number of
points to evaluate.
